---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Firstly the file **activity.zip** is unziped then the   
**activity.csv** file is loaded to R as a data.table.

```{r}
  library(data.table)
  unzip("activity.zip");
  data <- data.table(read.csv("activity.csv"));
```
##  What is mean total number of steps taken per day?
The following histogram shows the histogram of the number of  
steps taken each day.

```{r, cache = T}
# Sum steps by day
sumSteps <- data[, sum(steps, na.rm = TRUE), by = date];
# Display the result in a histogram
hist(sumSteps$V1, xlab = "Number of steps per day");
```

The following code shows the mean and median of the total number of steps taken per day.  

```{r, cache = TRUE}
  sumSteps[ , mean(V1, na.rm = TRUE)]
  sumSteps[ , median(V1, na.rm = TRUE)]

```

## What is the average daily activity pattern?

The folowing plot show the 5-minute interval average 
```{r, cache = TRUE}
averageDaily <- data[ , mean(steps, na.rm = TRUE), by = interval];
plot(x = averageDaily$interval, y = averageDaily$V1, xlab = "5-Minute Interval", 
     ylab = "Average");
```

The interval which has the maximum average is: 
```{r, cache = TRUE}
averageDaily[which.max(V1), ]
```

### Filling missing values

It can be seen that only the variable step has missing values and it has 2304 missing values.  
```{r, cache = TRUE}
  summary(data)
```

The following code fill missing value by the 5-minute interval mean

```{r, cache = TRUE}
  fill_missing <- function(step, inter) {
    ifelse(!is.na(step), result <- (step), 
           result <- averageDaily[averageDaily[ ,interval == inter]]$V1)
    result;
  }
  filledData <- data;
  filledData$steps <- mapply(fill_missing, filledData$steps, filledData$interval);
```

The following histogram shows the distrubition of the number of steps with missing values filled.

```{r, cache = TRUE}
  sumFilled <- filledData[ , sum(steps), by = date];
  hist(sumFilled$V1, breaks = 10);
```

The following code computes the mean and median of the data with missing values filled.
```{r, cache = TRUE}
  sumFilled[ , mean(V1)]
  sumFilled[ , median(V1)]
```

The mean and median values are higher. 

## Are there differences in activity patterns between weekdays and weekends?

```{r, cache = TRUE, out.height= 5}
  
  convert <- function(day) {
    day <- weekdays(day);
    ifelse(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
           result <- "weekday",
           result <- "weekend");
    
    result
  }
  filledData$date <- as.Date(filledData$date);
  filledData[ , day := sapply(filledData$date, convert)];
  meanDay <- filledData[ , mean(steps), by = c("interval", "day")];
  meanWeekday <- meanDay[meanDay[ , day == "weekday"]]
  meanWeekend <- meanDay[meanDay[ , day == "weekend"]]

  par(mfrow = c(2,1));
  plot(x = meanWeekday$interval, y = meanWeekday$V1, xlab = "interval", ylab = "Mean 5-minutes");
  plot(x = meanWeekend$interval, y = meanWeekend$V1, xlab = "interval", ylab = "Mean 5-minutes"); 
```

There exists a difference between week days and weekend days 
